---
title: "Introduction to smolts2bugs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to smolts2bugs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Introduction
The model used to estimate smolt production is a Bayesian model written in the BUGS-language
and is run in the Blackbox environment. The input for the model must be saved in a number of
_.odc_ files. Odc files is a binary format and cannot be (easily) be written from R. The 
format of the input is quite hard to create correctly by hand with large matrix and column
names that must be in a specific format to be recognized.

To help the process _smolts2bugs_ contains functions that can format capture/recapture data
and save an Excel-file where each sheet corresponds to one Blackbox input file. The Blackbox
.odc files can the be created by cut-n-paste between Excel and Blackbox.

## How to use Sötebasen data

The prefered dataflow is to first have quality assured data in Sötebasen. The
package Smoltreg have routines to check for errors, clean data and prepare a
file for Sötebasen import. All data in
Sötebasen with _AnstrTyp == "Smoltfälla"_ is regulary exported to CSV-files
that live in \\\\storage-dh.slu.se\\restricted$\\Sötebasen\\Exporter.

This package (_smolts2bugs_) comes with functions to read these files and format
mark/recapture data to an excel-file that is very close to the .ODC files that
BlackBox needs. The package also comes with a complete script that can be used
for the preparation of Sötebasen data to the Excelfile that can be used to
cut-n-paste data into Blackbox.

### Get the script
To get a copy of the script run:
```{r eval=FALSE}
# Save a script for the Sötebasen -> BlackBox conversion
smolts2bugs::get_script(scriptname = "söte2bugs.R")
# see ?smolts2bugs::get_script for other options
```

The recommended workflow is to load the script i Rstudio, adjust some variables
for your river, species and year and step through the code while checking that
you get reasonable data on each step. There are a couple of places in the code
where you need to make a decision on what you want to do.

How to choose what fish to include?
```{r eval=FALSE}
fish <- sdb_read_catch_recatch(Art = species, VattenNamn = river, Year = year) %>%
  filter(between(Längd1, minlength, maxlength) | is.na(Längd1))
#    filter(StadiumKod %in% c("S1", "S2", "S3") | Behandling == "Återfångad&utsatt")
```
This code reads all fish of one species for one river and year. Historically 
we have used the fish length to choose only fish between a min and max. But in
2023 in Ljungan we tried to use smoltification status, assesed by field personel,
to choose fish.

How to choose start and stop date?
In Sötebasen we get start and stop date for the trapping period but sometimes
we have long leading and/or trailing period with zero catches. It may be useful
to reduce the length of the period as the run time of the model seems to be very
dependant of the number of days.
```{r eval=FALSE}
# Uncomment the lines in this code and change the dates to reduce the length of
# study period.
dates <- dates %>%
#  mutate(AnstrDatumStart = as.Date("2023-05-11"),
#         AnstrDatumSlut = as.Date("2023-06-22")) %>% # Don't use long period of leading and trailing zero catch
  mutate(N_days = 1 + as.numeric(AnstrDatumSlut - AnstrDatumStart),
         start_day_of_year = as.POSIXlt(AnstrDatumStart)$yday,
         start_date = AnstrDatumStart,
         stop_date = AnstrDatumSlut) %>%
  select(N_days, start_day_of_year, start_date, stop_date)

```

The script will the assemble and format the data in BlackBox-format. Formated
data is saved in a separate folder under your current working directory (_getwd()_).
The script also saves RData_dump.Rdata with both data and meta data and a Rmarkdown
file (_results_river_species_year.Rmd). The rmarkdown file can procduce a human
readable version of the results when the moddel is run and results saved.


## Description of _smolts2bugs::format_Data2()_?

The bulk of the work is done by the function format_Data2() that will create the
matrix that should go into Data2_1group.odc. Step one is to create a data.frame
where each row represents one fish. The data.frame *must* contain three manatory
columns _capture_day_, _recapture_day_ and _marked_. Columns _capture_day_,
_recapture_day_ are the day number the fish was caught with the first day of the
experiment is 1. Column _marked_ is a logical indicating if the fish was marked
(subject for recapture) or not. The parameter _env_ is a data.frame with two
mandatory columns "water_temp" and "water_level". The number of rows in _env_
must be equal to parameter _ndays_ that indicates how many days the experiment
was running. The optional parameter _missing_days_ is a numerical vector of day
numbers where the trap wasn't operational.

Below is a small data set demonstrating an experiment with 5 fishing days where 
3 fish was caught, 2 of those were marked and both marked fish were recaptured.
The result is a matrix with dimensions ndays x ndays + 6.

```{r setup}
library(smolts2bugs)
ndays <- 5
(smolts <- data.frame(
   capture_day = c(2,2,3),
   recapture_day = c(NA, 3, 4),
   marked = c(FALSE, TRUE, TRUE)))
(envdata <- data.frame(
   w_temp = c(7, 7, 8, 8.5, 9),
   w_level = c(100, 110, 105, 150, 145)))

result <- format_Data2(fish = smolts, env = envdata, ndays = ndays)
result
```
The formated matrix should the be written to an Excel file using the function 
_save_bugsdata()_. This will create a file with 6 sheets _Data1_, _Data2_,
_Inits1c1_, _Inits1c2_, _Inits2c1_ and _Inits2c2_. Each sheet corresponds to a
Blackbox .odc file. By default the file will be saved in the current directory
and be named _bugsdata.xlsx_. Existing file will be overwritten.

```{r eval=FALSE}
save_bugsdata(result)
```

You can also save the data with R _save()_ with the function _save_Rdatadump()_,
this file can later be loaded with _load()_. Besides the smolt data
_save_Rdatadump()_ will save various meta data describes the experiment. The
data will be saved as a list named _rdata_. By default the file will be saved in
the current directory and be named _RData_dump.RData_. Existing file will be
overwritten.

```{r eval=FALSE}
save_Rdatadump(result, river = "Mörrumsån", species = "Lax",
               startd = "2020-03-15", stopd = "2020-05-20",
               missing_days = c(7, 15:18))
```




